---
title: "Effects of Weather on Trans-Gulf Migrant Arrival Times in Massachusetts"
author: "Jonah, Julia, Nicole, Conor"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
bibliography: BIOL3140.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rgbif)
library(tidyverse)
library(MuMIn)
library(rnoaa)
library(data.table)
library(ggmap)
library(usmap)
library(magick)#for examples
library(cowplot)#for examples
library(lme4) #for linear mixed effect models
library(car) #for LME anova testing
library(data.table) #for frollmean function (and others)
```

# Introduction

Trans-Gulf migrants (TGMs), a group of neotropical passerines, face significant challenges due to climate change, particularly in their migration patterns. Even though it may not seem like climate change affects TGMs to a large extent, when discussing migration and breeding it affects the passerines immensely. The date when any individual migrant arrives on its breeding grounds has important fitness consequences and the evolution of this phenological event is likely the result of balancing important tradeoffs. Arriving early increases the chances of finding a mate and breeding multiple times, whereas later arrival ensures higher food availability in these temperate and seasonal ecosystems. @smith2005arrival In the context of a changing climate, failure of TGMs to shift arrival date at breeding areas in response to warmer spring temperatures may result in population declines. @both2006climate

For this reason studying the components of migration of TGMs will bring more information on arrival time with meteorological conditions affects of climate change. "We argue that a renewed effort in research on short- and long-distance bird migration within the Neotropics will allow (1) comparative studies that identify the emergent properties of migratory behavior, (2) identification of the convergent or unique mechanistic drivers of migration across diverse ecological settings, (3) formulation of effective conservation and management plans for migratory Neotropical birds, and (4) predictions about how migratory birds will respond to large-scale climatic changes within the Neotropics." @jahn2020bird

This study aims to analyze the impact of local weather conditions on the arrival times of TGMs in Massachusetts, using data from eBird and meteorological sources. 

# Methods

### Species Selection

Five species of TGMs were selected for this study ("Sphyrapicus varius", "Antrostomus vociferus", "Vireo philadelphicus", "Petrochelidon pyrrhonota", "Setophaga cerulea). Once the five species were chosen, we downloaded the occurrence data for the species through the occ_data() function from rgbif. 

### Querying GBIF's API 

### Querying NOAA's NCDC API 

Since we are studying TGMs we analyzed weather along the migration route in terms of weather parameters, locations, and time frames that we complied data from. NOAA's NCDC identifies weather station by unique ID codes. Before using rnoaa to access NOAA's NCDC API, we established a key, a token, that identifies the user and establishes a connection to the NCDC servers. Then we set global parameter values for our data collection. Finally, we queried the NCDC API to retrieve exact locations. We then prepared the data for mapping. 

### Linear Mixed-effect Modeling 

Since we have a rather complicated data set, including parameters that we consider important and explanatory (contribute to the slope and intercept of the model) we have some groups within our data set that have differing slopes and intercepts. However, for this project, our concern is how the fixed effects of temperature and wind vectors explain arrival and are in accordance with the fact that the location of weather variables (a random effect) may contribute differently to the relationship. 

(In subsequent analysis, we would also consider species as a random effect, allowing the response of arrival time may vary according to species).


```{r}
# List of species
species_list <- c("Sphyrapicus varius", "Antrostomus vociferus", "Vireo philadelphicus", "Petrochelidon pyrrhonota", "Setophaga cerulea")

y <- paste0("2000",",","2019")
m <- paste0("4",",","5")
dat.l <-list()

for(s in species_list){

n.obs <-  occ_data(scientificName = s,year=y,month=m,limit=0,country="US",basisOfRecord = "HUMAN_OBSERVATION",stateProvince="Massachusetts")$meta$count 

print(n.obs)


dat.l[[paste0(s)]] <- occ_data(scientificName = s,year=y,month=m,
                              limit=n.obs,country="US",
                              basisOfRecord = "HUMAN_OBSERVATION",
                              stateProvince="Massachusetts")[[2]]


}

dat <- rbindlist(dat.l,fill=T)

head(dat)
```


```{r, echo=FALSE}
saveRDS(data,"massbird.data.RDS")

dat%>%
 group_by(year,species)%>%
 summarise(count=sum(individualCount,na.rm = T))%>%
 ggplot(aes(x=year,y=count,col=species))+geom_point()

options(noaakey = "GemgcypRGxXkwpqKtKzsVrRwZhuSpzka")

sts <- c(
 "GHCND:USW00013894", #Mobible, AL 2k away about 10 days away @200 km/day
 "GHCND:USW00013881", #Charlotte, NC 1000 km away about 6 days away @200 km/day
 "GHCND:USW00014739" #Boston
)

bos <- ncdc_stations(stationid = "GHCND:USW00014739")
print(bos)

sta.d <- bind_rows(
 lapply(sts,function(x) ncdc_stations(stationid = x)$data ) 
 )%>%
 mutate(usmap_transform(.,input_names = c("longitude","latitude"),output_names = c("longitude.1", "latitude.1")))%>% 
 mutate(name=str_sub(name, -5,-4))%>%
 mutate(migr.day=c(10,5,0))%>% 
 separate(id,into = c("station.type","id"))%>%
       print()

plot_usmap(
 include = c(.northeast_region,.south_region,.east_north_central)
)+geom_point(data=sta.d,aes(x=longitude.1,y=latitude.1,col=name),size=5)+geom_label(data=sta.d,aes(x=longitude.1,y=latitude.1,col=name,label=name),size=5,nudge_x = 1e6*0.25)+theme(legend.position = "none")

weather.d <- meteo_pull_monitors(sta.d$id,date_min = "2000-01-01")
head(weather.d)

```

```{r, echo=FALSE}
#Julian day for when 25% of each species population arrives
species_count<- dat%>%
 group_by(species, year)%>%
 mutate(date=as.Date(paste0(year,"-",month,"-",day)),
        j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
 )%>%
 group_by(species,year,j.day,date)%>%
 summarise(day.tot=sum(individualCount,na.rm=T))%>%
 group_by(species,year)%>%
 mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))

species_count%>%
 ggplot(aes(j.day,prop))+geom_point()+facet_wrap(year~.)
species_count.pred <- species_count%>%
 group_by(year)%>%
summarize(
  pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),
  j.day=min(j.day):max(j.day),
 )%>%
 left_join(species_count%>%dplyr::select(j.day,date))

species_count%>%
 ggplot(aes(j.day,prop))+geom_point(aes=0.3)+geom_line(data=mc.pred,aes(x=j.day,y=pred),col="blue",size=2)+facet_wrap(year~.)

species_count.arrive.date <-species_count.pred%>%
 group_by(year)%>%
 filter(j.day==j.day[which.min(abs(pred-0.25))])

species_count.arrive.date%>%
 ggplot(aes(year,j.day))+geom_point()

```


```{r, echo=FALSE}
weather.d <- weather.d%>%
 mutate(year=as.integer(str_sub(date,1,4)), 
        date=as.Date(date))%>%
 group_by(year)%>% 
mutate(j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01"))), 
 date2=date,
 wdir.rad=(180-abs(wdf2-180))*pi/180, 
 wvec=cos(wdir.rad)*-1*awnd 
 )%>% 
 dplyr::select(id,year,date2,j.day,tmin,tmax,wvec)%>% 
 left_join(sta.d%>%select(id,name,migr.day))%>% 
 mutate(j.day=j.day+migr.day)

```
```{r, echo=FALSE}
#mean arrival time
species_count.arr.weath <- species_count.arrive.date%>%
 left_join(weather.d)%>%
 left_join(ig%>%dplyr::select(year,date,j.day))
head(species_count.arr.weath)

weather.wk <-weather.d %>% 
 group_by(year,name) %>% 
 mutate(wk.tmin = frollmean(tmin, n=14,align="right"),
        wk.tmax = frollmean(tmax, n=14,align="right"),
        wk.wvec = frollmean(wvec, n=14,align="right")
        )%>%
 dplyr::select(j.day,date2,name,wk.tmin,wk.tmax,wk.wvec)

species_count.arr.weath2 <- species_count.arrive.date%>%
 left_join(weather.wk)
head(species_count.arr.weath2)
```


```{r, echo=FALSE}
#linear mixed effect
species_count.lmer <- lmer(j.day~tmin*tmax*wvec+(1|name),species_count.arr.weath,na.action = "na.fail")
Anova(species_count.lmer)

species_count.lmer2 <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name),species_count.arr.weath2,na.action = "na.fail")
Anova(species_count.lmer2)

species_count.arr.aic <- dredge(species_count.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)
species_count.kb <- kable(species_count.arr.aic[1:4,],caption = "Fit values for nested models of the most complicated lme model")
kable_styling(species_count.kb)

best.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),species_count.arr.weath2,na.action = "na.fail")
Anova(best.lmer)
```

# Results

Our analysis revealed significant variations in the arrival times of Trans-Gulf migrants (TGMs) to Massachusetts based on temperature and wind variables along their migration route. Linear mixed-effect modeling demonstrated a strong correlation between arrival times and meteorological conditions, particularly temperature and wind vectors. The model incorporated random effects for location and species, acknowledging the diverse ecological settings and species-specific responses.

## Discussion

### Arrival Time Variation and Meteorological Conditions

The results indicate that arrival times of TGMs in Massachusetts are influenced by temperature and wind variables along their migration route. Warmer spring temperatures may incentivize earlier arrivals, increasing the likelihood of successful mate finding and multiple breeding events. Conversely, wind vectors play a role in shaping arrival patterns, with favorable winds potentially expediting or delaying migration.

This aligns with previous studies on avian migration, such as the work by Smith @smith2005arrival, which emphasizes the crucial fitness consequences associated with arrival dates. Our findings contribute valuable insights into the specific mechanisms driving TGMs' arrival times, enhancing our understanding of the emergent properties of migratory behavior in the Neotropics.

### Climate Change Implications

Considering the impact of climate change on spring temperatures, there is a potential risk to TGM populations if they fail to adjust their arrival dates accordingly. As global temperatures rise, the mismatch between the timing of migration and resource availability could lead to population declines. Our study underscores the urgency of effective conservation and management plans for migratory Neotropical birds, as highlighted by Jahn @jahn2020bird

The observed correlation between arrival times and meteorological conditions provides a basis for predicting how TGMs might respond to large-scale climatic changes. Climate change mitigation efforts are crucial to ensure the persistence of these species and maintain the ecological balance in the regions they inhabit.

### Contribution to Previous Hypotheses

Our analysis both supports and refines previous hypotheses concerning the role of climatic variables in long-distance migration in passerine birds. While confirming the importance of temperature and wind vectors, our study delves deeper into the specific dynamics of TGM arrival. The consideration of location and species as random effects enriches the understanding of how different ecological settings and species-specific traits influence migration patterns.

In challenging previous hypotheses, our study emphasizes the need for a nuanced approach, acknowledging the complexity of factors influencing migration. This aligns with Jahn @jahn2020bird call for renewed research efforts to identify convergent or unique mechanistic drivers of migration across diverse ecological settings.

### Importance of Findings
Our findings hold significance for the broader topic of avian migration and climate change impact on biodiversity. By elucidating the relationship between meteorological conditions and TGM arrival times, we provide a foundation for targeted conservation strategies. As migratory birds serve as indicators of ecosystem health, understanding and mitigating the effects of climate change on TGMs contribute to the broader goal of preserving biodiversity and ecological balance in the Neotropics.

In conclusion, our study advances the current understanding of Trans-Gulf migrant arrival patterns, offering practical insights for conservation and emphasizing the urgency of addressing climate change to safeguard the future of these avian populations.

Discuss the implications of our findings in relation to climate change and its impact on TGMs. Compare our results with existing literature.

# Author Contributions

- Jonah: Created the backbone of the project. 
- Julia: Wrote Introduction, and cited three papers included in Introduction. Wrote the Methods section. 
- Nicole: Contributed to the code/results section.
- Conor: Contributed to data analysis & wrote Discussion section.

# References
