---
title: "Effects of Weather on Trans-Gulf Migrant Arrival Times in Massachusetts"
author: "Jonah, Julia, Nicole, Conor"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
bibliography: BIOL3140.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rgbif)
library(tidyverse)
library(MuMIn)
library(rnoaa)
library(data.table)
library(ggmap)
library(usmap)
library(magick)#for examples
library(cowplot)#for examples
library(lme4) #for linear mixed effect models
library(car) #for LME anova testing
library(data.table) #for frollmean function (and others)
```

# Introduction
Trans-Gulf migrants (TGMs), a group of neotropical passerines, face significant challenges due to climate change, particularly in their migration patterns. Even though it may not seem like climate change affects TGMs to a large extent, when discussing migration and breeding it affects the passerines immensely. The date when any individual migrant arrives on its breeding grounds has important fitness consequences and the evolution of this phenological event is likely the result of balancing important tradeoffs. Arriving early increases the chances of finding a mate and breeding multiple times, whereas later arrival ensures higher food availability in these temperate and seasonal ecosystems. @smith2005arrival In the context of a changing climate, failure of TGMs to shift arrival date at breeding areas in response to warmer spring temperatures may result in population declines. @both2006climate

For this reason studying the components of migration of TGMs will bring more information on arrival time with meteorological conditions affects of climate change. "We argue that a renewed effort in research on short- and long-distance bird migration within the Neotropics will allow (1) comparative studies that identify the emergent properties of migratory behavior, (2) identification of the convergent or unique mechanistic drivers of migration across diverse ecological settings, (3) formulation of effective conservation and management plans for migratory Neotropical birds, and (4) predictions about how migratory birds will respond to large-scale climatic changes within the Neotropics." @jahn2020bird

This study aims to analyze the impact of local weather conditions on the arrival times of TGMs in Massachusetts, using data from eBird and meteorological sources. 

# Methods

### Species Selection

Five species of TGMs ("Sphyrapicus varius", "Antrostomus vociferus", "Vireo philadelphicus", "Petrochelidon pyrrhonota", "Setophaga cerulea) were selected for this study. Once the five species were chosen, we downloaded the occurrence data for the species through the occ_data() function from rgbif. 

```{r}
# List of species
species_list <- c("Sphyrapicus varius", "Antrostomus vociferus", "Vireo philadelphicus", "Petrochelidon pyrrhonota", "Setophaga cerulea")

# Data compilation and analysis loop
dat_l <- list()
models_lmer <- list()
models_lmer2 <- list()

for (s in species_list) {
  
  # Occurrence data retrieval
  n_obs <- occ_data(scientificName = s, year = paste0("2000", ",", "2019"), month = paste0("4", ",", "5"), limit = 0, country = "US", basisOfRecord = "HUMAN_OBSERVATION", stateProvince = "Massachusetts")$meta$count
  
  dat_l[[s]] <- occ_data(scientificName = s, year = paste0("2000", ",", "2019"), month = paste0("4", ",", "5"), limit = n_obs, country = "US", basisOfRecord = "HUMAN_OBSERVATION", stateProvince = "Massachusetts")[[2]]
  
  # Logistic modeling
  mc <- dat_l[[s]] %>%
    group_by(year) %>%
    mutate(
      date = as.Date(paste0(year, "-", month, "-", day)),
      j.day = julian(date, origin = as.Date(paste0(unique(year), "-01-01")))
    ) %>%
    group_by(species, year, j.day, date) %>%
    summarise(day.tot = sum(individualCount, na.rm = TRUE)) %>%
    group_by(species, year) %>%
    mutate(prop = cumsum(day.tot / sum(day.tot, na.rm = TRUE))) %>%
    filter(year > 1999)
  
  mc.pred <- mc %>%
    group_by(year) %>%
    summarize(
      pred = predict(nls(prop ~ SSlogis(j.day, Asym, xmid, scal)), newdata = data.frame(j.day = min(mc$j.day):max(mc$j.day))),
      j.day = min(mc$j.day):max(mc$j.day)
    ) %>%
    left_join(mc %>% dplyr::select(j.day, date))
  
  # Weather data operations
  weather.d <- meteo_pull_monitors(sta.d$id, date_min = "2000-01-01")
  mc.arr.weath <- mc.arrive.date %>%
    left_join(weather.d) %>%
    left_join(mc %>% dplyr::select(year, date, j.day))
  
  weather.wk <- weather.d %>%
    group_by(year, name) %>%
    mutate(
      wk.tmin = frollmean(tmin, n = 14, align = "right"),
      wk.tmax = frollmean(tmax, n = 14, align = "right"),
      wk.wvec = frollmean(wvec, n = 14, align = "right")
    ) %>%
    dplyr::select(j.day, date2, name, wk.tmin, wk.tmax, wk.wvec)
  
  mc.arr.weath2 <- mc.arrive.date %>%
    left_join(weather.wk)
  
  # Linear mixed-effect modeling
  lmer_model <- lmer(j.day ~ wk.tmin * wk.tmax * wk.wvec + (1|name), mc.arr.weath2, na.action = "na.fail")
  lmer_model2 <- lmer(j.day ~ wk.tmin + wk.tmax + wk.wvec + (1|name), mc.arr.weath2, na.action = "na.fail")
  
  models_lmer[[s]] <- lmer_model
  models_lmer2[[s]] <- lmer_model2
}

# Model testing and Anova testing
aic_list <- list()

for (i in seq_along(species_list)) {
  mc.arr.aic <- dredge(models_lmer2[[i]], fixed = c("wk.tmin", "wk.tmax", "wk.wvec"))
  aic_list[[species_list[i]]] <- mc.arr.aic
}

# Display AIC values in a table
for (i in seq_along(species_list)) {
  cat("\nAIC values for", species_list[i], ":\n")
  print(kable(aic_list[[species_list[i]]][1:4, ], caption = paste("Fit values for nested models of", species_list[i], "lme model")))
}
```


# Results

```{r}
# List of species
species_list <- c("Sphyrapicus varius", "Antrostomus vociferus", "Vireo philadelphicus", "Petrochelidon pyrrhonota", "Setophaga cerulea")

y <- paste0("2000",",","2019")
m <- paste0("4",",","5")
dat.l <-list()

for(s in species){

n.obs <-  occ_data(scientificName = s,year=y,month=m,limit=0,country="US",basisOfRecord = "HUMAN_OBSERVATION",stateProvince="Massachusetts")$meta$count 

print(n.obs)


dat.l[[paste0(s)]] <- occ_data(scientificName = s,year=y,month=m,
                              limit=n.obs,country="US",
                              basisOfRecord = "HUMAN_OBSERVATION",
                              stateProvince="Massachusetts")[[2]]


}

dat <- rbindlist(dat.l,fill=T)

head(dat)
```


```{r, echo=FALSE}
saveRDS(data,"massbird.data.RDS")

dat%>%
 group_by(year,species)%>%
 summarise(count=sum(individualCount,na.rm = T))%>%
 ggplot(aes(x=year,y=count,col=species))+geom_point()

options(noaakey = "GemgcypRGxXkwpqKtKzsVrRwZhuSpzka")

sts <- c(
 "GHCND:USW00013894", #Mobible, AL 2k away about 10 days away @200 km/day
 "GHCND:USW00013881", #Charlotte, NC 1000 km away about 6 days away @200 km/day
 "GHCND:USW00014739" #Boston
)

bos <- ncdc_stations(stationid = "GHCND:USW00014739")
print(bos)

sta.d <- bind_rows(
 lapply(sts,function(x) ncdc_stations(stationid = x)$data ) 
 )%>%
 mutate(usmap_transform(.,input_names = c("longitude","latitude"),output_names = c("longitude.1", "latitude.1")))%>% 
 mutate(name=str_sub(name, -5,-4))%>%
 mutate(migr.day=c(10,5,0))%>% 
 separate(id,into = c("station.type","id"))%>%
       print()

plot_usmap(
 include = c(.northeast_region,.south_region,.east_north_central)
)+geom_point(data=sta.d,aes(x=longitude.1,y=latitude.1,col=name),size=5)+geom_label(data=sta.d,aes(x=longitude.1,y=latitude.1,col=name,label=name),size=5,nudge_x = 1e6*0.25)+theme(legend.position = "none")

weather.d <- meteo_pull_monitors(sta.d$id,date_min = "2000-01-01")
head(weather.d)

```

```{r, echo=FALSE}
#Julian day for when 25% of each species population arrives
species_count<- dat%>%
 group_by(species, year)%>%
 mutate(date=as.Date(paste0(year,"-",month,"-",day)),
        j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
 )%>%
 group_by(species,year,j.day,date)%>%
 summarise(day.tot=sum(individualCount,na.rm=T))%>%
 group_by(species,year)%>%
 mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))

species_count%>%
 ggplot(aes(j.day,prop))+geom_point()+facet_wrap(year~.)
species_count.pred <- species_count%>%
 group_by(year)%>%
summarize(
  pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),
  j.day=min(j.day):max(j.day),
 )%>%
 left_join(species_count%>%dplyr::select(j.day,date))

species_count%>%
 ggplot(aes(j.day,prop))+geom_point(aes=0.3)+geom_line(data=mc.pred,aes(x=j.day,y=pred),col="blue",size=2)+facet_wrap(year~.)

species_count.arrive.date <-species_count.pred%>%
 group_by(year)%>%
 filter(j.day==j.day[which.min(abs(pred-0.25))])

species_count.arrive.date%>%
 ggplot(aes(year,j.day))+geom_point()

```


```{r, echo=FALSE}
weather.d <- weather.d%>%
 mutate(year=as.integer(str_sub(date,1,4)), 
        date=as.Date(date))%>%
 group_by(year)%>% 
mutate(j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01"))), 
 date2=date,
 wdir.rad=(180-abs(wdf2-180))*pi/180, 
 wvec=cos(wdir.rad)*-1*awnd 
 )%>% 
 dplyr::select(id,year,date2,j.day,tmin,tmax,wvec)%>% 
 left_join(sta.d%>%select(id,name,migr.day))%>% 
 mutate(j.day=j.day+migr.day)

```
```{r, echo=FALSE}
#mean arrival time
species_count.arr.weath <- species_count.arrive.date%>%
 left_join(weather.d)%>%
 left_join(ig%>%dplyr::select(year,date,j.day))
head(species_count.arr.weath)

weather.wk <-weather.d %>% 
 group_by(year,name) %>% 
 mutate(wk.tmin = frollmean(tmin, n=14,align="right"),
        wk.tmax = frollmean(tmax, n=14,align="right"),
        wk.wvec = frollmean(wvec, n=14,align="right")
        )%>%
 dplyr::select(j.day,date2,name,wk.tmin,wk.tmax,wk.wvec)

species_count.arr.weath2 <- species_count.arrive.date%>%
 left_join(weather.wk)
head(species_count.arr.weath2)
```


```{r, echo=FALSE}
#linear mixed effect
species_count.lmer <- lmer(j.day~tmin*tmax*wvec+(1|name),species_count.arr.weath,na.action = "na.fail")
Anova(species_count.lmer)

species_count.lmer2 <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name),species_count.arr.weath2,na.action = "na.fail")
Anova(species_count.lmer2)

species_count.arr.aic <- dredge(species_count.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)
species_count.kb <- kable(species_count.arr.aic[1:4,],caption = "Fit values for nested models of the most complicated lme model")
kable_styling(species_count.kb)

best.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),species_count.arr.weath2,na.action = "na.fail")
Anova(best.lmer)
```

# Discussion

Discuss the implications of our findings in relation to climate change and its impact on TGMs. Compare our results with existing literature.

# Author Contributions

- Jonah
- Julia: Wrote Introduction, and cited three papers included in Introduction. Wrote the Methods section. 
- Nicole
- Conor: Contributed to data analysis & wrote Discussion section.

# References





